---
title: "exercisePrediction.Rmd"
author: "Tinniam V Ganesh"
date: "Wednesday, May 20, 2015"
output: html_document
---
# Executive Summary 
This project performs a logistic model fit of the exercise measurements taken from wearables. 

Read the data
```{r}
library(dplyr)
library(caret)
library(rattle)
library(rpart.plot)
training <- read.csv("pml-training.csv",stringsAsFactors=FALSE)
testing <- read.csv("pml-testing.csv",stringsAsFactors=FALSE)
```

#Explore the data
```{r}
# Check the column names
str(training)
```

# Clean the data. Remove columns which have a NAs or empty spaces for most rows. Select all columns for which all rows (16922) have valid values.

```{r cache=TRUE}
col <- NULL
for( i in 1: length(names(training))){
   a <- training[,i] != "" & !is.na(training[,i])
   b <- sum(a)
   
   #Check if the number of values in column is complete
   if(b == nrow(training) ) {
      col <- c(col,i)
   }
  
}
# Subset all colums which have data
train <- select(training,col)
```

## Drop all columns that don't have relevance to the outcome
```{r cache=TRUE}
#Drop columns that don't have relevance
train <-select(train, -new_window)
train <- select(train, -user_name)
train <- select(train, -cvtd_timestamp)
train <- select(train, -X)
train <- select(train,-raw_timestamp_part_1)
train <- select(train,-raw_timestamp_part_2)
train <- select(train,-contains("total"))
train <- select(train,-num_window)

```


# Normalize all the variables 
Subtract the Mean and divide by the Standard Deviation
``` {r cache=TRUE}
# Mean center and divide by SD to normalize all columns
value <- length(names(train)) -1
for( i in 1: value ){
  train[,i] <- (train[,i] - mean(train[,i]))/sd(train[,i])
}
```

# Convert the outcome 'classe' variable to a factor
```{r cache=TRUE}
train$classe <- as.factor(train$classe)
```


#  See the details of the train set
```{r}
dim(train)
```

# Use logistic regression to make the prediction for the training set
```{r cache=TRUE}
modFit <- train(classe ~ ., method="rpart",data= train)
#summary(modFit)
```

# Plot the classification with plot and fancyRpartPlot
```{r classfication-chart, cache=TRUE}
plot(modFit$finalModel,uniform=TRUE,main='Classification based on exercises') 
text(modFit$finalModel,use.n=TRUE,all=TRUE,cex=0.8)
```

```{r classification-fancy, cache=TRUE}
# Use the fancy Rpart plot for a better visualization of the classification
fancyRpartPlot(modFit$finalModel)
```

# Subset the same columns and remove irrelevant columns  from the test set
```{r cache=TRUE}
test <- select(testing,col)

test <-select(test, -new_window)
test <- select(test, -user_name)
test <- select(test, -cvtd_timestamp)
test <- select(test, -X)
test <- select(test,-raw_timestamp_part_1)
test <- select(test,-raw_timestamp_part_2)
test <- select(test,-contains("total"))
test <- select(test,-num_window)
```

# Predict with the test set
```{r cache=TRUE}
predictions <- predict(modFit,newdata=test)
predictions

```

# Conclusion: 
1. The training and the test set data were read from the CSV files
2. The data was cleaned based on the following criteria 
   a. For columns which have blanks remove elements. 
   b. For columns for which there are NA values remove the elements
   c. Remove all irrelevant columns e.g user_name, run_window which don't contribute to the outcome
3. A classification CART model was used for the outcone 'classe' variable
4. After cleaning the total number of predictors is 48
5. A classification chart and a fancy Rpart plot is created
6. The testing set is also cleaned like the training set
7. Finally the predictions for the test set is computed and shown